Fig2: Candidate genes

### 2020-04-17 analysis/figure-2.Rmd
new candidate gene/GO analysis

---
title: "Fig3"
author: "Rayna M Harris"
date: "3/26/2020"
output: md_document
---

# all things box plots

## candidate genes

```{r}
library(tidyverse)
library(ggtext)
library(cowplot)
library(ggpubr)

@@

knitr::opts_chunk$set(echo = TRUE, message = F, fig.path = "../figures/")

hormonepapergenes <- c("GNRH1", "CGNRH-R", "GNRHR",  
                       "FSHB", "FSHR", "LHCGR", # no LSH
                       "VIP", "VIPR1",
                       "ESR1", "AR", "CYP19A1", "HSD17B1", # receptors synthesis
                       "PGR" , "HSD3B2", 
                       "PRL", "PRLR")

curleychampagnegenes <- c("AVPR1A", "ADRA2A",    
                         "COMT", "CRH", "CRHBP", "CRHR1", "CRHR2", 
                          "DRD1", "DRD4","ESR1", "ESR2", #ERa ERbeta
                          "FOS", "HTR2C", "MEST", "NR3C1", #GR
                          "OPRM1", "PGR", "PRL", "PRLR",  "SLC6A4" #5HTT
                         )

parentalcareGO <- c("AVP", "AVPR1A", "BRINP1", "CREBRF", "DBH", "DRD1",  
                    "GNAQ", "HUBB", "KALRN", "MBD2","NPAS1", "NPAS3",
                     "NR3C1", "OPRK1", "OXT", "OXTR", "PRL", "PTEN")

# can't find NOS1 or OXTR or FOX1B or HTR5A
```

@@

```{r }
geneids <- read_csv("../metadata/00_geneinfo.csv") %>% select(-X1)

@@

hypvsd <- getcandidatevsd(curleychampagnegenes, "hypothalamus", sexlevels)
pitvsd <- getcandidatevsd(curleychampagnegenes, "pituitary", sexlevels)
gonvsd <- getcandidatevsd(curleychampagnegenes, "gonad", sexlevels)
head(hypvsd)
tail(gonvsd)

candidatevsd <- rbind(hypvsd, pitvsd)
candidatevsd <- rbind(candidatevsd, gonvsd)

unique(candidatevsd$gene)
```

## Figs

```{r }
# summary DEG results from DESeq2
candidateDEGS <- read_csv("../../musicalgenes/data/allDEG.csv") %>%
  select(-X1) %>%
  filter(gene %in% curleychampagnegenes)
head(candidateDEGS)

# get genes to plot
candidateDEGS %>% filter(tissue == "hypothalamus", sex == "female", comparison != "control_bldg") 
candidateDEGS %>% filter(tissue == "hypothalamus", sex == "male", comparison != "control_bldg") 
candidateDEGS %>% filter(tissue == "pituitary", sex == "male", comparison != "control_bldg") 
candidateDEGS %>% filter(tissue == "pituitary", sex == "female", comparison != "control_bldg") 
candidateDEGS %>% filter(tissue == "gonad", sex == "female", comparison != "control_bldg")

# summarize all differences
candidateDEGtable <- candidateDEGS %>%
  mutate(posneg = ifelse(lfc >= 0, "+", "-")) %>%
  mutate(geneposneg = paste(gene, posneg, sep = "")) %>%
  select(sex, tissue, comparison, geneposneg) %>%
  group_by(sex, tissue, comparison) %>%
   arrange(geneposneg) %>%
  summarize(genes = str_c(geneposneg, collapse = " ")) %>% 
  pivot_wider(names_from = comparison, values_from = genes)  %>% 
  ungroup(tissue) %>%
  mutate(tissue = factor(tissue, levels =  tissuelevel))  %>% 
  arrange(sex, tissue)  %>% 
  select(sex, tissue, bldg_lay, lay_inc.d3, inc.d3_inc.d9,  
         inc.d9_inc.d17,  hatch_n5, control_bldg)
candidateDEGtable
```

```{r fig3, fig.width = 7}

@@

```{r writefiles}
#write.csv(candidatevsd, "../../musicalgenes/data/candidatecounts.csv")
#write.csv(candidatevsd, "../results/candidatecounts.csv")
write.csv(candidateDEGtable, "../results/candidateDEGtable.csv")

```

>>>>

---
title: "Fig2: Candidate genes"
author: "Rayna M Harris"
date: "3/26/2020"
output: md_document
---

# Candidate gene analysis

```{r}
library(tidyverse)
library(ggtext)
library(cowplot)
library(ggpubr)
library(knitr)
library(kableExtra)

@@

knitr::opts_chunk$set(echo = TRUE, message = F, fig.path = "../figures/")
```

## Candidate Genes (from literature and relevant GO terms)

```{r GO}

# all genes in this parental care dataset

geneids <- read_csv("../metadata/00_geneinfo.csv") %>%
  select(-X1) %>%
  rename("NCBI" = "entrezid",
         "gene" = "Name")

# From Ch 17 Evolution of parental care
curleychampagnegenes <- c("AVPR1A", "ADRA2A",    
                         "COMT", "CRH", "CRHBP", "CRHR1", "CRHR2", 
                          "DRD1", "DRD4","ESR1", "ESR2", #ERa ERbeta
                          "FOS", "HTR2C", "MEST", "NR3C1", #GR
                          "OPRM1", "PGR", "PRL", "PRLR",  "SLC6A4") #5HTT
curleychampagnegenes <- as.data.frame(curleychampagnegenes) %>%
  mutate(GO = "parentalcare", gene = curleychampagnegenes)  %>%
  select(GO, gene)
curleychampagnegenes

# Candidate GO terms 
GO_path <- "../metadata/goterms/"   # path to the data
GO_files <- dir(GO_path, pattern = "*.txt") # get file names
GO_pathfiles <- paste0(GO_path, GO_files)

GOgenesLong <- GO_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_table(.x, col_types = cols(), col_names = FALSE), .id = "file_name") %>% 
  mutate(GO = sapply(strsplit(as.character(file_name),'../metadata/goterms/'), "[", 2)) %>% 
  mutate(GO = sapply(strsplit(as.character(GO),'.txt'), "[", 1)) %>% 
  mutate(gene = sapply(strsplit(as.character(X1), "[\\\\]|[^[:print:]]" ), "[", 2)) %>% 
  select(GO, gene)  %>%
  filter(gene != "Symbol") %>%
  distinct(GO,gene)  %>%
  mutate(gene = toupper(gene)) %>%
  rbind(., curleychampagnegenes) %>%
  arrange(gene) %>%
  left_join(., geneids, by = "gene") %>%
  drop_na() 
GOgenesLong

GOgenesWide <- GOgenesLong %>% 
  pivot_wider(
    names_from = GO,
    values_from = gene) %>%
  mutate(numGOs = 4 - rowSums(is.na(.))) %>%
  arrange(desc(numGOs)) %>%
  select(-numGOs) %>%
  left_join(geneids, by = c("geneid","NCBI"))
GOgenesWide

# genes I can't find in dataset 
# NOS1 or OXTR or FOX1B or HTR5A
```

## Candidate DEGs

```{r DEGs}

candidategenes <- GOgenesLong %>% distinct(gene) %>% pull(gene)

# summary DEG results from DESeq2
candidateDEGS <- read_csv("../results/suppletable1.csv") %>%
  filter(gene %in% candidategenes) %>%
  mutate(posneg = ifelse(lfc >= 0, "+", "-"),
         sex = recode(sex, "female" = "F", "male" = "M" ),
         tissue = recode(tissue, 
                         "hypothalamus" = "H",
                         "pituitary" = "P", "gonad" = "G")) %>%
  mutate(res = paste(sex, tissue, posneg, sep = "")) %>%
  select(gene, res, comparison)  %>%
  group_by(gene,  comparison) %>%
  summarize(res = str_c(res, collapse = " ")) %>%
  pivot_wider(names_from = comparison, values_from = res) %>%
  select(gene, bldg_lay, lay_inc.d3, inc.d3_inc.d9, inc.d9_inc.d17, hatch_n5, n5_n9)
candidateDEGS

table1 <- left_join(candidateDEGS, GOgenesWide) %>%
  select(gene, bldg_lay:n5_n9, parentalcare, parentalbehavior, NCBI, ) %>%
  mutate(parentalcare = if_else(is.na(parentalcare), " ", "X"),
         parentalbehavior = if_else(is.na(parentalbehavior), " ", "X")) %>%
  rename("Literature" = "parentalcare", "GO" =  "parentalbehavior")
table1[is.na(table1)] <- " "
table1

kable(table1)
```

@@

```{r eval = F}

@@

head(allvsd)
candidategenes <- GOgenesLong %>% distinct(gene) %>% pull(gene)

hypvsd <- getcandidatevsd(candidategenes, "hypothalamus", sexlevels)
pitvsd <- getcandidatevsd(candidategenes, "pituitary", sexlevels)
gonvsd <- getcandidatevsd(candidategenes, "gonad", sexlevels)
candidatevsd <- rbind(hypvsd, pitvsd)
candidatevsd <- rbind(candidatevsd, gonvsd)
head(candidatevsd)
```

```{r fig2, fig.width = 7, eval = F}

@@

```{r writefiles}
#write.csv(candidatevsd, "../../musicalgenes/data/candidatecounts.csv")
#write.csv(candidatevsd, "../results/candidatecounts.csv")
write.csv(table1, "../results/table1.csv")
```
